---
title: "Reproducible Science Assignment"
output:
  html_document: default
  pdf_document: default
date: "2023-10-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## QUESTION 01: Data Visualisation for Science Communication

### a) Provide your figure here:

```{r, results='hide', message=FALSE, echo=FALSE}
 
#Load required packages for Q1
library(tidyverse)
library(palmerpenguins)
library(janitor)
library(here)
library(ggplot2)
```

```{r bad figure code, echo=FALSE}

# There is no need to provide the code for your bad figure, just use echo=FALSE so the code is hidden. Make sure your figure is visible after you knit it. 

#Clean data for easier data handling
badpenguins_clean <- penguins_raw %>%
  select(-Comments) %>%
  select(-starts_with("Delta")) %>%
     mutate(Species = case_when(
      Species == "Adelie Penguin (Pygoscelis adeliae)" ~ "Adelie",
      Species == "Chinstrap penguin (Pygoscelis antarctica)" ~ "Chinstrap",
      Species == "Gentoo penguin (Pygoscelis papua)" ~ "Gentoo")) %>%
  na.omit()


#Split up data by species
BadAdelie <- filter(badpenguins_clean, Species == "Adelie Penguin (Pygoscelis adeliae)")

BadChinstrap <- filter(badpenguins_clean, Species == "Chinstrap penguin (Pygoscelis antarctica)")

BadGentoo <- filter(badpenguins_clean, Species == "Gentoo penguin (Pygoscelis papua)")

#Make a data frame of the median body mass for each species of penguin
Baddata <- data.frame(Species = c("Adelie", "Chinstrap", "Gentoo"), Average = c(median(BadAdelie$`Body Mass (g)`), median(BadChinstrap$`Body Mass (g)`), median(BadGentoo$`Body Mass (g)`)))


#Define colours
badspecies_colours <- c("Adelie" = "green4",
                     "Chinstrap" = "gold4",
                     "Gentoo" = "cyan")

worsespecies_colours <- c("Adelie" = "gold4",
                     "Chinstrap" = "green4",
                     "Gentoo" = "cyan")



#Create plot
BadBar <- ggplot(data = badpenguins_clean,
       aes(x = Species, y = `Body Mass (g)`)) +
  geom_bar(stat = "summary", fun = "mean", aes(fill = Species)) +
  geom_jitter(aes(color = Species),
              cex = 6,
              position = position_jitter(
                seed = 21,
                width = 0.5)) +
  coord_cartesian(xlim = c(1, NA), ylim = c(3000, 5500)) +
  scale_fill_manual(values = badspecies_colours) +
  scale_color_manual(values = worsespecies_colours) +
  labs(title = "Plot of Mean and Individual Penguin Masses") +
  theme_classic()


#Show plot
BadBar


```

### b) Write about how your design choices mislead the reader about the underlying data (200-300 words).

------------------------------------------------------------------------

## QUESTION 2: Data Pipeline

### Introduction

```{r Library Setup, results='hide', message=FALSE}

#Loads libraries from renv file, such that the exact versions of the packages used initally are also used when re-running the code.
renv::restore()

#Loads the required packages
library(tidyverse)
library(palmerpenguins)
library(janitor)
library(here)
library(ggplot2)
library(svglite)
library(gridExtra)
library(patchwork)

#Loads the "PenguinFunctions" file within the "functions" folder, containing other functions that are used.
source(here("functions", "PenguinFunctions.R"))

```

```{r Data Exploration}

#Checks to see if user has "data" folder in the current working directory. If "no", creates the file so that the data files that will be created later are stored inside it, helping keep the directory organised. If "yes", notifies user that it does exist.
ifelse(file.exists(here("data")), print("data folder exists"), 
       dir.create(here("data")) )

#Creates the raw data file within the "data" folder that has now been produced.
write.csv(penguins_raw, here("data","penguins_raw.csv")) 

#Cleans the data by standardising headings, removing columns, shortening species names, and removing NA rows to avoid errors.
penguins_clean <- clean.penguin.data(penguins_raw)

#Saves the penguins_clean data as its own file, within the "data" folder, such that it can be viewed.
write.csv(penguins_clean, here("data", "penguins_clean.csv"))

#Sets the colours of the different penguin species, different colours to allow users to distinguish between penguin species.
species_colours <- c("Adelie" = "darkorange", 
                     "Chinstrap" = "purple", 
                     "Gentoo" = "cyan4")

#Sets what shape each point should be for each penguin species, to enable further distinction between different penguin species.
species_shape <- c("Adelie" = 15, 
                   "Chinstrap" = 16, 
                   "Gentoo" = 17)

#Plots a scatterplot as an exploratory figure, using the colours and shapes defined before, of the 3 species of bird, based on culmen length on the x-axis and culmen depth on the y-axis.
exploratory_fig <- ggplot(
  data = penguins_clean,
  aes(x = culmen_length_mm, 
      y = culmen_depth_mm)) +
  geom_point(aes(color = species, shape = species), cex = 1.2) +
  scale_color_manual(values = species_colours) +
  scale_shape_manual(values = species_shape) +
  labs(x = "Culmen length (mm)", 
       y = "Culmen depth (mm)",
       title = "A plot of culmen length against culmen depth of three penguin species",) +
  theme_bw()

#Show the scatterplot
exploratory_fig

#Like before, checks to see if a "figures" folder exists within the current directory. If no, it creates it, such that all figures can be stored in the same place.
ifelse(file.exists(here("figures")), print("figures folder exists"), 
       dir.create(here("figures")))

#Saves the exploratory figure as a .svg file, which keeps a high resolution when viewed.
save.as.svg(exploratory_fig, "figures/Penguins_exploratory.svg", 15, 15, 2)

```

### Hypothesis

The null hypothesis is that there is no correlation between culmen length and culmen depth within penguin species.

The alternative hypothesis is that there is a correlation between these measurements within the penguin species.

### Statistical Methods

The data were separated based on penguin species, and each species plotted on its own plot. The data were plotted as histograms to see if they followed a normal distribution, to determine whether a parametric or a non-parametric test would be appropriate. These histograms are saved as an additional figure. For adelie and gentoo penguins, both culmen length and culmen depth appears to be sufficiently consistent with a normal distribution to use parametric tests. For chinstrap penguins, consistency of the culmen lengths with the normal distribution was found after a Shapiro-Wilk test provided a sufficiently large P-value (0.1941). This was done as the histogram did not show a clear normal distribution. The Pearson's r correlation coefficient was used to characterise the relationship between variables of all three species.

```{r Statistics}

#Creates 3 sets of data based on the species of penguin involved, by removing all the rows that are not the species of interest (i.e: In "Adelie_data", removes all rows that are not "Adelie" in the "species" column).
Adelie_data <- filter(penguins_clean, species == "Adelie")

Chinstrap_data <- filter(penguins_clean, species == "Chinstrap")

Gentoo_data <- filter(penguins_clean, species == "Gentoo")

#Saves all Adelie points as a scatter plot to see any correlation
Adelie_plot <- ggplot(
  data = Adelie_data,
  aes(x = culmen_length_mm, 
      y = culmen_depth_mm)) +
  geom_point(aes(color = species, shape = species), cex = 1.2, 
             show.legend = FALSE) +
  scale_color_manual(values = species_colours) +
  scale_shape_manual(values = species_shape) +
  labs(x = "Culmen length (mm)", 
       y = "Culmen depth (mm)",
       title = "Adelie") +
  theme_bw()

#Saves all Chinstrap points as a scatter plot
Chinstrap_plot <- ggplot(
  data = Chinstrap_data,
  aes(x = culmen_length_mm, 
      y = culmen_depth_mm)) +
  geom_point(aes(color = species, shape = species), cex = 1.2,
             show.legend = FALSE) +
  scale_color_manual(values = species_colours) +
  scale_shape_manual(values = species_shape) +
  labs(x = "Culmen length (mm)", 
       y = "Culmen depth (mm)",
       title = "Chinstrap") +
  theme_bw()

#Saves all Gentoo points as a scatter plot
Gentoo_plot <- ggplot(
  data = Gentoo_data,
  aes(x = culmen_length_mm, 
      y = culmen_depth_mm)) +
  geom_point(aes(color = species, shape = species), cex = 1.2,
             show.legend = FALSE) +
  scale_color_manual(values = species_colours) +
  scale_shape_manual(values = species_shape) +
  labs(x = "Culmen length (mm)", 
       y = "Culmen depth (mm)",
       title = "Gentoo") +
  theme_bw()

#Plots the culmen lengths as a histogram to check for a normal distribution, with each histogram colour-coded according to species, using the same colour scheme seen in the exploratory figure for consistency.
Adelie_length_hist <- ggplot(data = Adelie_data, aes(x = culmen_length_mm)) + 
  geom_histogram(binwidth = 1, fill = "orange") +
  labs(x = "Culmen length (mm)", 
       y = "Frequency", 
       title = "Adelie Culmen Lengths") +
  theme_bw()
Chinstrap_length_hist <- ggplot(data = Chinstrap_data, aes(x = culmen_length_mm)) +
    geom_histogram(binwidth = 1, fill = "purple") +
  labs(x = "Culmen length (mm)", 
       y = "Frequency", 
       title = "Chinstrap Culmen Lengths") +
  theme_bw()
Gentoo_length_hist <- ggplot(data = Gentoo_data, aes(x = culmen_length_mm)) +
    geom_histogram(binwidth = 1, fill = "cyan4") +
  labs(x = "Culmen length (mm)", 
       y = "Frequency",
       title = "Gentoo Culmen Lengths") +
  theme_bw()

#Plots the culmen depths as a histogram to check for a normal distribution, once again labelling and colour-coding according to species.
Adelie_depth_hist <- ggplot(data = Adelie_data, aes(x = culmen_depth_mm)) + 
  geom_histogram(binwidth = 0.5, fill = "orange") +
  labs(x = "Culmen depth (mm)", 
       y = "Frequency",
       title = "Adelie Culmen Depths") +
  theme_bw()
Chinstrap_depth_hist <- ggplot(data = Chinstrap_data, aes(x = culmen_depth_mm)) +
    geom_histogram(binwidth = 0.5, fill = "purple") +
  labs(x = "Culmen depth (mm)", 
       y = "Frequency",
       title = "Chinstrap Culmen Depths") +
  theme_bw()
Gentoo_depth_hist <- ggplot(data = Gentoo_data, aes(x = culmen_depth_mm)) +
    geom_histogram(binwidth = 0.5, fill = "cyan4") +
  labs(x = "Culmen depth (mm)", 
       y = "Frequency",
       title = "Gentoo Culmen Depths") +
  theme_bw()

#Combines all histograms into a single figure
All_hists <- (Adelie_length_hist | Adelie_depth_hist) / (Chinstrap_length_hist | Chinstrap_depth_hist) / (Gentoo_length_hist | Gentoo_depth_hist)

#Save the histograms as a figure
save.as.svg(All_hists, "figures/Penguins_histograms.svg", 15, 15, 2)

#Show histograms
All_hists

#Shapiro-Wilk test to test for normality in chinstrap culmen lengths.
shapiro.test(Chinstrap_data$culmen_length_mm)

#The Pearson's correlation test is used. Outputs are saved as a "_test" object, allowing extraction of the outputs at a later stage.
Adelie_test <- cor.test(~ culmen_length_mm + culmen_depth_mm, data = Adelie_data)

Chinstrap_test <- cor.test(~ culmen_length_mm + culmen_depth_mm, 
                           data = Chinstrap_data)

Gentoo_test <- cor.test(~ culmen_length_mm + culmen_depth_mm, data = Gentoo_data)

```

### Results & Discussion

```{r Plotting Results}

#Creates an output table - a data frame of the T-values, Degrees of Freedom (Df), P.values, and #Correlation Coefficients (r) taken from the correlation tests:
Output_table <- data.frame(
  Species = c("Adelie", "Chinstrap", "Gentoo"),
  T.value = c(Adelie_test$statistic, Chinstrap_test$statistic,
              Gentoo_test$statistic),
  Df = c(Adelie_test$parameter, Chinstrap_test$parameter, Gentoo_test$parameter),
  P.value = c(Adelie_test$p.value, Chinstrap_test$p.value, Gentoo_test$p.value),
  Correlation.Coefficient = c(Adelie_test$estimate, Chinstrap_test$estimate,
                              Gentoo_test$estimate))

#Save the output table as a format that can be displayed visually
StatTable <- tableGrob(Output_table)

#This combines the 3 scatterplots of each species with the output table of the statistics
Combined <- (Adelie_plot | Chinstrap_plot | Gentoo_plot) / (StatTable)

#Saves the results figure in the "figures" folder
save.as.svg(Combined, "figures/Penguins_results.svg", 15, 10, 2)

#Show results figure
Combined
```

### Conclusion

There is a significant positive correlation between culmen length and culmen depth within all three species of penguins sampled in the Palmer penguins dataset. The null hypothesis is rejected in all three species.

**References:**

Gorman, K. B. et al. (2014) Ecological sexual dimorphism and environmental variability within a community of Antarctic penguins (Genus Pygoscelis). *PloS one*. [Online] 9 (3).

------------------------------------------------------------------------

## QUESTION 3: Open Science

### a) GitHub

*GitHub link:* <https://github.com/BatWeek/PenguinProjectAssessment>

### b) Share your repo with a partner, download, and try to run their data pipeline.

*Partner's GitHub link:*

*You **must** provide this so I can verify there is no plagiarism between you and your partner.*

### c) Reflect on your experience running their code. (300-500 words)

-   *What elements of your partner's code helped you to understand their data pipeline?*

-   *Did it run? Did you need to fix anything?*

-   *What suggestions would you make for improving their code to make it more understandable or reproducible, and why?*

-   *If you needed to alter your partner's figure using their code, do you think that would be easy or difficult, and why?*

### d) Reflect on your own code based on your experience with your partner's code and their review of yours. (300-500 words)

-   *What improvements did they suggest, and do you agree?*

-   *What did you learn about writing code for other people?*
